---
title: "Data Wrangling"
author: "Derrick Yam & Michael Lopez"
date: "December, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Required libraries

```{r}
library(readr)
library(dplyr) 
library(splines)
library(ggplot2)
library(randomForest)
library(mgcv)
```

## Read in data

*Note*: `plays` and `games` data from Armchair Analysis, while `ratingswinprob` is from Football Outsiders

```{r}
plays <- read_csv("Data/PLAY.csv")
games <- read_csv("Data/GAME.csv")
ratingswinprob <- read.csv("Data/ratings.csv", stringsAsFactors = FALSE)

```

## Data cleaning

Steps include

- create new variables
- filter to regular season games, quarters 1-4, from 2004 onwards

```{r}
#Ensure names always line up
plays[plays$off == "LA",]$off <- "STL"
plays[plays$def == "LA",]$def <- "STL"
games[games$h == "LA",]$h <- "STL"
games[games$v == "LA",]$v <- "STL"

#Armchair note: swap offense and Defense when a team punts.
plays.new <- plays %>%
  mutate(team.temp = off,
         off = ifelse(dwn == 4 & type == "PUNT", def, off),
         def = ifelse(dwn == 4 & type == "PUNT", team.temp, def)) %>% 
  mutate(points.temp = ptso,
         ptso = ifelse(dwn == 4 & type == "PUNT", ptsd, ptso),
         ptsd = ifelse(dwn == 4 & type == "PUNT", points.temp, ptsd)) %>% 
  mutate(ptsdiff = ptso-ptsd,
         seconds = 15*60*(4-qtr) + (60*min) + sec,
         score.lev = ptsdiff/sqrt(seconds + 1),
         totalpoints = ptso + ptsd) 

#Join with variables from the game data file
NFLwin <- left_join(plays.new, games)

#Create a win variable for the win probability models
NFLwin <- NFLwin %>% 
  mutate(Home = ifelse(off == h, 1, 0),
         won.home = ifelse(ptsh > ptsv, 1, 0),
         won.off = ifelse(Home == 1 & won.home == 1 |
                          Home == 0 & won.home == 0, 1, 0))

#Filter out observations we will not use in our win probability model or analyses.
NFLwin <- NFLwin %>% 
  filter(qtr <5) %>% 
  filter(seas > 2003) %>% 
  filter(type != "NOPL") %>% 
  filter(wk < 18)

#Clean ratings data
names(ratingswinprob)[1] <- "off"
names(ratingswinprob)[4] <- "seas"
names(ratingswinprob)[5] <- "def"

ratings.off <- dplyr::select(ratingswinprob, 1:4)
ratings.def <- dplyr::select(ratingswinprob, 4:7)

#Merge ratings with NFL data
NFLwin <- NFLwin %>%
  left_join(ratings.off) %>%
  left_join(ratings.def)

NFLwin$won.home <- as.numeric(NFLwin$won.home)
NFLwin$won.off <- as.numeric(NFLwin$won.off)

##Impute Missing Data for weather conditions in Dome games
NFLwin[is.na(NFLwin$wspd),]$wspd<-0
NFLwin[is.na(NFLwin$humd),]$humd<-60
NFLwin[is.na(NFLwin$temp),]$temp<-70
```

## Construct win probability model framework 

Create training and test data frames

```{r}
#Set seed for random samples and random forests
set.seed(1)

#Define the games to be in the test data set
games.test <- games %>%
  filter(wk < 18) %>%
  group_by(seas, wk) %>%
  sample_n(2, replace = FALSE)

#Define the test data set for the rf win probability model
NFLwin.test <- NFLwin %>% 
  filter(gid %in% games.test$gid)

##Training data set
NFLwin.train <- NFLwin %>% 
  filter(!gid %in% games.test$gid)

#Check the dimensions to make sure we didn't lose any observations
dim(NFLwin)
dim(NFLwin.test)
dim(NFLwin.train)

# Strip the training dataset of extra variables
NFLwin.train <- NFLwin.train %>% 
   select(won.off, sprv , ptsdiff , seconds , score.lev , totalpoints,
                yfog , dwn , ytg , timo , timd)

str(NFLwin.train)

##Omit Observations with missing Values

NFLwin.train <- na.omit(NFLwin.train)
str(NFLwin.train) 
#Note: won.off needs to be an integer for the randomForest to run properly 
```

## Random Forest Win Probability Model

Implement random forest win probability model of Lock and Nettleton (JQAS, 2014). This model is available on Github.  

```{r}
NFLwinprob.Lock.reg = get(load("Models/winprob_lock_rf.RData"))

print(NFLwinprob.Lock.reg)
NFLwinprob.Lock.reg$importance #Show variable importance


##Probability of winning
NFLwin.test$wp.Lock <- predict(NFLwinprob.Lock.reg, NFLwin.test)

##Probability split into categories
NFLwin.test$predict.cat <- cut(NFLwin.test$wp.Lock, 20)

##Compare those win probability categories with the true mean outcomes within that group.
NFLwin.test <- mutate(NFLwin.test, won.off = as.numeric(as.character(won.off)))

##Create a data frame to plot calibration
agg.Lock.reg <- NFLwin.test %>%
  group_by(predict.cat) %>%
  summarize(mean.win = mean(won.off), n.game = n()) %>%
  rename(predict.cat = predict.cat) 

estWP <- 0:19/20 + 0.025
WPaccuracy <- cbind(agg.Lock.reg[1:20,], estWP)

#View calibration
p.Lock <- ggplot(agg.Lock.reg, aes(predict.cat, mean.win))
p.Lock + geom_point() + 
  geom_abline(intercept = -0.025, slope = 1/20) + 
  scale_x_discrete(name = "Estimated Win Probability",
                   breaks=c("(0.000106,0.051]", "(0.201,0.251]","(0.45,0.5]","(0.7,0.75]", "(0.95,1]"), 
                   labels=c("0.00", "0.25",".5",".75", "1.00")) + 
  ylab("Proportion of Games Won") + 
  ggtitle("Accuracy of the Lock win probability model") + 
  theme_bw() + 
  theme(legend.position = "none")

#Predict win probability for all plays after we confirmed the accuracy on the test data set
NFLwin$wp.Lock <- predict(NFLwinprob.Lock.reg, NFLwin)

# Apply outcomes from random forest model to observations

#Define the home team's win probability, since the RF gives the offensive wp
NFLwin <- NFLwin %>% 
  mutate(wp.home.Lock = ifelse(Home == 1, wp.Lock,  1 - wp.Lock))

#Calculate the change in win probability from each play for the home team
NFLwin <- NFLwin %>% 
  group_by(gid) %>% 
  mutate(wp.home.lead.Lock = lead(as.numeric(wp.home.Lock), 1)) %>% 
  mutate(deltawp.home.Lock = wp.home.lead.Lock - wp.home.Lock) 

#Calculate the change in win probability from each play for the offensive team
NFLwin <- NFLwin %>% 
  mutate(wp.off.lead.Lock = ifelse(Home == 1, wp.home.lead.Lock, 1 - wp.home.lead.Lock)) %>% 
  mutate(deltawp.off.Lock = wp.off.lead.Lock - wp.Lock)
```



## Generalized Additive Win Probability Model

Implement version of generalized additive model used by Horowitz et al with `nflscrapr` package in R.  This model is available on Github. 

```{r}
NFLwinprob.Horowitz.reg = get(load("Models/winprob_horowitz_gam.RData"))

# Run the prediction function on our Test data frame
NFLwin.test$wp.Scrapr <- predict.gam(NFLwinprob.Horowitz.reg, NFLwin.test, type = "response")

##Probability split into categories
NFLwin.test$predict.cat.NFLS <- cut(NFLwin.test$wp.Scrapr, breaks = 0:20/20)

agg.NFLS.reg <- NFLwin.test %>%
  group_by(predict.cat.NFLS) %>%
  summarize(mean.win.s = mean(won.off), n.game.s = n()) %>%
  rename(predict.cat.NFLS = predict.cat.NFLS) 

estWP <- 0:19/20 + 0.025
WPaccuracy <- cbind(agg.NFLS.reg[1:20,], estWP)

p.NFLS <- ggplot(agg.NFLS.reg, aes(predict.cat.NFLS, mean.win.s))
p.NFLS + geom_point() + 
  geom_abline(intercept = -0.025, slope = 1/20) + 
  scale_x_discrete(name = "Estimated Win Probability",
                   breaks=c("(0.000106,0.051]", "(0.201,0.251]","(0.45,0.5]","(0.7,0.75]", "(0.95,1]"), 
                   labels=c("0.00", "0.25",".5",".75", "1.00")) + 
  ylab("Proportion of Games Won") + 
  ggtitle("Accuracy of the Scrapr win probability model") + 
  theme_bw() + 
  theme(legend.position = "none")


## NFLscrapr Applying Outcomes to Observations

NFLwin$wp.Scrapr <- predict.gam(NFLwinprob.Horowitz.reg, NFLwin, type = "response")

NFLwin <- mutate(NFLwin,  wp.home.Scrapr = ifelse(Home == 1, wp.Scrapr,  1 - wp.Scrapr))

NFLwin <- NFLwin %>% 
  group_by(gid) %>% 
  mutate(wp.home.lead.Scrapr = lead(as.numeric(wp.home.Scrapr), 1)) %>% 
  mutate(deltawp.home.Scrapr = wp.home.lead.Scrapr - wp.home.Scrapr) 

NFLwin <- NFLwin %>% 
  mutate(wp.off.lead.Scrapr = ifelse(Home == 1, wp.home.lead.Scrapr, 1 - wp.home.lead.Scrapr)) %>% 
  mutate(deltawp.off.Scrapr = wp.off.lead.Scrapr - wp.Scrapr) 
```

## Average the two win probabilities and view accuracy plot together

Here, we make an accuracy plot of each win probability model using predictions on the test data.  

```{r}
plot.all <- NFLwin.test %>% 
  select(gid, pid, wp.Scrapr, wp.Lock, won.off) %>% 
  mutate(scrapr.cat = cut_number(wp.Scrapr, 20), 
         lock.cat = cut_number(wp.Lock, 20))

lock.probs <- plot.all %>% group_by(lock.cat) %>% 
  summarise(estimated = mean(wp.Lock), observed = mean(won.off), type = "Lock")
scrapr.probs <- plot.all %>% group_by(scrapr.cat) %>% 
  summarise(estimated = mean(wp.Scrapr), observed = mean(won.off), type = "nflscrapR")
wp.all <- bind_rows(lock.probs, scrapr.probs)



p.wpacc <- ggplot(wp.all, aes(x = estimated, y = observed, shape = type, colour = type))
wpacc <- p.wpacc + 
  geom_point() + 
  geom_abline(intercept = 0, slope = 1) + 
  scale_x_continuous(labels = scales::percent, "Estimated win rate") + 
  scale_y_continuous(labels = scales::percent, "Observed win rate") + 
  scale_colour_manual(values = c("black", "darkgrey"),
                                  name  ="Win Probability Model",
                            breaks=c("Lock", "nflscrapR"),
                            labels=c("Lock", "nflscrapR")) +
  scale_shape_manual(values = c(19, 3),name  ="Win Probability Model",
                           breaks=c("Lock", "nflscrapR"),
                            labels=c("Lock", "nflscrapR")) + 
  ylab("Proportion of Games Won") + 
  ggtitle("Accuracy of the win probability models") + 
  theme_bw() + 
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = c(.85, .15))
wpacc
#ggsave(wpacc, file = "Figures/WP_accuracy_plot.png", height = 6, width = 8)
```


## Define NYT 4th Down Bot Strategy

```{r}
df.goforit <- expand.grid(yfog = 1:99, distance = 1:10)
df.goforit <- df.goforit %>%
  mutate(coaches.do = "Punt", coaches.should = "Punt")


df.goforit <- df.goforit %>%
  mutate(
  coaches.do = ifelse((yfog >=55 & yfog <= 84 & distance == 1), "Go for it", coaches.do),
  coaches.do = ifelse((yfog >=84 & yfog <= 91 & distance == 1), "FG", coaches.do),
  coaches.do = ifelse((yfog > 91 & distance == 1), "Go for it", coaches.do),
  coaches.do = ifelse((yfog > 60 & yfog <=68 & distance ==2), "Go for it", coaches.do),
  coaches.do = ifelse((yfog > 68 & distance ==2), "FG", coaches.do),
  coaches.do = ifelse((yfog > 63 & yfog <=67 & distance ==3), "Go for it", coaches.do),
  coaches.do = ifelse((yfog > 67 & distance ==3), "FG", coaches.do),
  coaches.do = ifelse((yfog > 64 & yfog <=66 & distance ==4), "Go for it", coaches.do),
  coaches.do = ifelse((yfog > 66 & distance ==4), "FG", coaches.do),
  coaches.do = ifelse((yfog > 63 & yfog <=65 & distance ==5), "Go for it", coaches.do),
  coaches.do = ifelse((yfog > 65 & distance ==5), "FG", coaches.do),
  coaches.do = ifelse((yfog > 65 & distance ==6), "FG", coaches.do),
  coaches.do = ifelse((yfog > 65 & distance ==7), "FG", coaches.do),
  coaches.do = ifelse((yfog > 64 & distance ==8), "FG", coaches.do),
  coaches.do = ifelse((yfog > 66 & distance ==9), "FG", coaches.do),
  coaches.do = ifelse((yfog > 66 & distance ==10), "FG", coaches.do)
  )

df.goforit <- df.goforit %>%
  mutate(
   coaches.should = ifelse((distance == 1), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 28 & distance == 2), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 41 & yfog <=80 & distance == 3), "Go for it", coaches.should), 
   coaches.should = ifelse((yfog > 80 & yfog <=96 & distance == 3), "FG", coaches.should), 
   coaches.should = ifelse((yfog > 96 & distance == 3), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 45 & yfog <=72 & distance == 4), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 72 & distance == 4), "FG", coaches.should),
   coaches.should = ifelse((yfog > 50 & yfog <=68 & distance == 5), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 68 & distance == 5), "FG", coaches.should),
   coaches.should = ifelse((yfog > 53 & yfog <= 67 & distance == 6), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 67 & distance == 6), "FG", coaches.should),
   coaches.should = ifelse((yfog > 59 & yfog <= 65 & distance == 7), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 65 & distance == 7), "FG", coaches.should),
   coaches.should = ifelse((yfog > 59 & yfog <=64 & distance == 8), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 64 & distance == 8), "FG", coaches.should),
   coaches.should = ifelse((yfog > 62 & yfog <=65 & distance == 9), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 65 & distance == 9), "FG", coaches.should),
   coaches.should = ifelse((yfog > 63 & distance == 10), "FG", coaches.should)
   )

df.goforit <- df.goforit %>% 
  filter(yfog + distance <= 100 & yfog + distance >= 10) %>% 
  rename(ytg = distance) %>% 
  mutate(dwn = 4)

table(df.goforit$coaches.do, df.goforit$coaches.should)

```

## More filtering and cleaning

```{r}
## Data check -- do we have 8 regular season home games for each team?
NFLwin %>% 
  group_by(gid) %>% 
  slice(1) %>% 
  ungroup() %>% 
  group_by(seas, h) %>% 
  count() %>% 
  print.data.frame()

### Did the team go for it on 4th down?
NFLwin <- NFLwin %>% 
  mutate(Attempt.4 = ifelse(dwn == 4 & type == "RUSH" |  dwn == 4 & type == "PASS", 1, 0))

### Filter to 4th down plays only, re-define time
NFLwin.fourth <- NFLwin %>%
   filter(dwn==4) %>% 
   mutate(time = floor((3600 - seconds)/60))

#Weather Variables
Precipitation <- c("Chance Rain",  "Flurries", "Light Rain", 
                   "Light Showers", "Light Snow", "Rain", "Snow", "Thunderstorms")
Dome <- c("Closed Roof", "Covered Roof", "Dome")


NFLwin.fourth$cond.cat <- "DRY"
NFLwin.fourth[NFLwin.fourth$cond %in% Precipitation,]$cond.cat <- "Precipitation"
NFLwin.fourth[NFLwin.fourth$cond %in% Dome,]$cond.cat <- "Dome"

NFLwin.fourth$cond.cat <- factor(NFLwin.fourth$cond.cat, levels = c("DRY", "Dome",  "Precipitation"))

##Redefine Point Differential as a Categorical Variable and then as an ordered factor
NFLwin.fourth$pointdiff <- cut(NFLwin.fourth$ptsdiff,
                     c(-Inf, -17, -9, -4, -1, .1, 3, 8, 16, Inf),
                     labels=c("pdiff.M4", "pdiff.M3",  "pdiff.M2", "pdiff.M1", 
                              "pdiff.T", "pdiff.P1", "pdiff.P2", "pdiff.P3", "pdiff.P4"))

NFLwin.fourth$pointdiff <- factor(NFLwin.fourth$pointdiff, 
                                  levels = c("pdiff.M4", "pdiff.M3",  "pdiff.M2", "pdiff.M1", 
                                             "pdiff.T", "pdiff.P1", "pdiff.P2", "pdiff.P3", "pdiff.P4"))

## Create the final set of fourth down plays
NFL <- inner_join(NFLwin.fourth, df.goforit) %>% ungroup()
NFL <- NFL[!is.na(NFL$deltawp.off.Scrapr),] 
NFL <- NFL %>% filter(pid != 588564)  #win probability calculation is off in the Lock model for this play (no time left, 15 point game)

dim(NFL)
table(NFL$Attempt.4)
table(NFL$coaches.should)
table(NFL$Attempt.4, NFL$coaches.should)

NFL %>% 
  filter(coaches.should == "Go for it") %>% 
  group_by(gid, off) %>% 
  count() %>% 
  group_by(n) %>% 
  count()


#Export the data set into a new csv file. to be used in matching
write.csv(NFL, file = "Data/NFL_fourthdown_cleaned.csv", row.names = FALSE)
```










