---
title: "Data Wrangling"
author: "Derrick Yam & Michael Lopez"
date: "December, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Required libraries

```{r}
library(readr)
library(dplyr) 
library(splines)
library(ggplot2)
library(randomForest)
library(mgcv)
```

## Read in data

*Note*: `plays` and `games` data from Armchair Analysis, while `ratingswinprob` is from Football Outsiders

```{r}
#plays <- read_csv("Data/PLAY.csv")
#games <- read_csv("Data/GAME.csv")
#ratingswinprob <- read.csv("Data/ratings.csv", stringsAsFactors = FALSE)
#
```

## Data cleaning

Steps include

- create new variables
- filter to regular season games, quarters 1-4, from 2004 onwards

```{r}
#Ensure names always line up
#The only franchise that changed during these seasons was the Rams to LA
plays[plays$off == "LA",]$off <- "STL"
plays[plays$def == "LA",]$def <- "STL"
games[games$h == "LA",]$h <- "STL"
games[games$v == "LA",]$v <- "STL"

#Armchair note: swap offense and Defense when a team punts.
plays.new <- plays %>%
  mutate(team.temp = off,
         off = ifelse(dwn == 4 & type == "PUNT", def, off),
         def = ifelse(dwn == 4 & type == "PUNT", team.temp, def)) %>% 
  mutate(points.temp = ptso,
         ptso = ifelse(dwn == 4 & type == "PUNT", ptsd, ptso),
         ptsd = ifelse(dwn == 4 & type == "PUNT", points.temp, ptsd)) %>% 
  mutate(ptsdiff = ptso-ptsd,
         seconds = 15*60*(4-qtr) + (60*min) + sec,
         score.lev = ptsdiff/sqrt(seconds + 1),
         totalpoints = ptso + ptsd) 

#Join with variables from the game data file
NFLwin <- left_join(plays.new, games)

#Create a win variable for the win probability models
NFLwin <- NFLwin %>% 
  mutate(Home = ifelse(off == h, 1, 0),
         won.home = ifelse(ptsh > ptsv, 1, 0),
         won.off = ifelse(Home == 1 & won.home == 1 |
                          Home == 0 & won.home == 0, 1, 0))

#Filter out observations we will not use in our win probability model or analyses.
NFLwin <- NFLwin %>% 
  filter(qtr < 5) %>% 
  filter(seas > 2003) %>% 
  filter(type != "NOPL") %>% 
  filter(wk < 18)

#Clean ratings data
names(ratingswinprob)[1] <- "off"
names(ratingswinprob)[4] <- "seas"
names(ratingswinprob)[5] <- "def"

ratings.off <- dplyr::select(ratingswinprob, 1:4)
ratings.def <- dplyr::select(ratingswinprob, 4:7)

#Merge ratings with NFL data
NFLwin <- NFLwin %>%
  left_join(ratings.off) %>%
  left_join(ratings.def)

NFLwin$won.home <- as.numeric(NFLwin$won.home)
NFLwin$won.off <- as.numeric(NFLwin$won.off)

##Impute Missing Data for weather conditions in Dome games
NFLwin[is.na(NFLwin$wspd),]$wspd<-0
NFLwin[is.na(NFLwin$humd),]$humd<-60
NFLwin[is.na(NFLwin$temp),]$temp<-70
```

## Construct win probability model framework 

Create training and test data frames

```{r}
#Set seed for random samples and random forests
set.seed(1)

#Define the games to be in the test data set
games.test <- games %>%
  filter(wk < 18) %>%
  group_by(seas, wk) %>%
  sample_n(2, replace = FALSE)

#Define the test data set for the rf win probability model
NFLwin.test <- NFLwin %>% 
  filter(gid %in% games.test$gid)

##Training data set
NFLwin.train <- NFLwin %>% 
  filter(!gid %in% games.test$gid)

#Check the dimensions to make sure we didn't lose any observations
dim(NFLwin)
dim(NFLwin.test)
dim(NFLwin.train)

isTRUE(dim(NFLwin)[1] == dim(NFLwin.test)[1] + dim(NFLwin.train)[1])

# Strip the training dataset of extra variables
NFLwin.train <- NFLwin.train %>% 
   select(won.off, sprv , ptsdiff , seconds , score.lev , totalpoints,
                yfog , dwn , ytg , timo , timd)

str(NFLwin.train)

##Omit Observations with missing Values

NFLwin.train <- na.omit(NFLwin.train)
str(NFLwin.train) 
#Note: won.off needs to be an integer for the randomForest to run properly 
```

## Random Forest Win Probability Model

Implement random forest win probability model of Lock and Nettleton (JQAS, 2014). This model is available on Github. This requires use of the Armchair Analysis data.

```{r}
#Note: this win probability model is trained strictly on our training data set.
NFLwinprob.Lock.reg = get(load("Models/winprob_lock_rf.RData"))

print(NFLwinprob.Lock.reg)
NFLwinprob.Lock.reg$importance #Show variable importance (mostly out of curiousity)

#Apply Lock win probability for all plays 
NFLwin$wp.Lock <- predict(NFLwinprob.Lock.reg, NFLwin)

#Define the win probability for the home team
NFLwin <- NFLwin %>% 
  mutate(wp.home.Lock = ifelse(Home == 1, wp.Lock,  1 - wp.Lock))

#Create a variable for the change in win probability using the RF model.
NFLwin <- NFLwin %>% 
  group_by(gid) %>% 
  mutate(wp.home.lead.Lock = lead(as.numeric(wp.home.Lock), 1)) %>% 
  mutate(deltawp.home.Lock = wp.home.lead.Lock - wp.home.Lock) 

NFLwin <- NFLwin %>% 
  mutate(wp.off.lead.Lock = ifelse(Home == 1, wp.home.lead.Lock, 1 - wp.home.lead.Lock)) %>% 
  mutate(deltawp.off.Lock = wp.off.lead.Lock - wp.Lock)
```



## Generalized Additive Win Probability Model

Implement version of generalized additive model used by Horowitz et al with `nflscrapr` package in R.  This model is available on Github. This requires use of the Armchair Analysis data. 

```{r}
#Note: again this is trained only on our training data set.
NFLwinprob.Horowitz.reg = get(load("Models/winprob_horowitz_gam.RData"))

## NFLscrapr for all plays
NFLwin$wp.Scrapr <- predict.gam(NFLwinprob.Horowitz.reg, NFLwin, type = "response")
NFLwin <- mutate(NFLwin,  wp.home.Scrapr = ifelse(Home == 1, wp.Scrapr,  1 - wp.Scrapr))

##Create a variable for the change in win probability using the GAM
NFLwin <- NFLwin %>% 
  group_by(gid) %>% 
  mutate(wp.home.lead.Scrapr = lead(as.numeric(wp.home.Scrapr), 1)) %>% 
  mutate(deltawp.home.Scrapr = wp.home.lead.Scrapr - wp.home.Scrapr) 

NFLwin <- NFLwin %>% 
  mutate(wp.off.lead.Scrapr = ifelse(Home == 1, wp.home.lead.Scrapr, 1 - wp.home.lead.Scrapr)) %>% 
  mutate(deltawp.off.Scrapr = wp.off.lead.Scrapr - wp.Scrapr) 
```

## Average the two win probabilities and view accuracy plot together

Here, we make an accuracy plot of each win probability model using predictions on the test data.  

```{r}
## Predict WP on test data
NFLwin.test$wp.Lock <- predict(NFLwinprob.Lock.reg, NFLwin.test)
NFLwin.test$wp.Scrapr <- predict.gam(NFLwinprob.Horowitz.reg, NFLwin.test, type = "response")

##Create a data frame for the WP accuracy plot
plot.all <- NFLwin.test %>% 
  select(gid, pid, wp.Scrapr, wp.Lock, won.off) %>% 
  mutate(scrapr.cat = cut_number(wp.Scrapr, 20), 
         lock.cat = cut_number(wp.Lock, 20))

lock.probs <- plot.all %>% 
  group_by(lock.cat) %>% 
  summarise(estimated = mean(wp.Lock), observed = mean(won.off), type = "Lock")

scrapr.probs <- plot.all %>%
  group_by(scrapr.cat) %>% 
  summarise(estimated = mean(wp.Scrapr), observed = mean(won.off), type = "nflscrapR")

wp.all <- bind_rows(lock.probs, scrapr.probs)

p.wpacc <- ggplot(wp.all, aes(x = estimated, y = observed,
                              shape = type, colour = type))

wpacc <- p.wpacc + 
  geom_point(size = 2.5) + 
  geom_abline(intercept = 0, slope = 1) + 
  scale_x_continuous(labels = scales::percent, "Estimated win rate") + 
  scale_y_continuous(labels = scales::percent, "Observed win rate") + 
  scale_colour_manual(values = c("black", "red"),
                                  name  ="Win Probability Model",
                            breaks=c("Lock", "nflscrapR"),
                            labels=c("Lock", "nflscrapR")) +
  scale_shape_manual(values = c(19, 3), name  ="Win Probability Model",
                           breaks=c("Lock", "nflscrapR"),
                            labels=c("Lock", "nflscrapR")) + 
  ylab("Proportion of Games Won") + 
  ggtitle("Accuracy of the win probability models") + 
  theme(plot.title = element_text(hjust = 0.5, size = rel(1.2)),
        legend.position = c(.85, .15),
        axis.title.y = element_text(size = rel(1.2)),
        axis.title.x = element_text(size = rel(1.2)))
wpacc
#ggsave(wpacc, file = "Figures/WP_accuracy_plot.png", height = 6, width = 8)
```


## Define NYT 4th Down Bot Strategy

```{r}
df.goforit <- expand.grid(yfog = 1:99, ytg = 1:10)
df.goforit <- df.goforit %>%
  mutate(coaches.do = "Punt", coaches.should = "Punt")


df.goforit <- df.goforit %>%
  mutate(
  coaches.do = ifelse((yfog >=55 & yfog <= 84 & ytg == 1), "Go for it", coaches.do),
  coaches.do = ifelse((yfog >=84 & yfog <= 91 & ytg == 1), "FG", coaches.do),
  coaches.do = ifelse((yfog > 91 & ytg == 1), "Go for it", coaches.do),
  coaches.do = ifelse((yfog > 60 & yfog <=68 & ytg ==2), "Go for it", coaches.do),
  coaches.do = ifelse((yfog > 68 & ytg ==2), "FG", coaches.do),
  coaches.do = ifelse((yfog > 63 & yfog <=67 & ytg ==3), "Go for it", coaches.do),
  coaches.do = ifelse((yfog > 67 & ytg ==3), "FG", coaches.do),
  coaches.do = ifelse((yfog > 64 & yfog <=66 & ytg ==4), "Go for it", coaches.do),
  coaches.do = ifelse((yfog > 66 & ytg ==4), "FG", coaches.do),
  coaches.do = ifelse((yfog > 63 & yfog <=65 & ytg ==5), "Go for it", coaches.do),
  coaches.do = ifelse((yfog > 65 & ytg ==5), "FG", coaches.do),
  coaches.do = ifelse((yfog > 65 & ytg ==6), "FG", coaches.do),
  coaches.do = ifelse((yfog > 65 & ytg ==7), "FG", coaches.do),
  coaches.do = ifelse((yfog > 64 & ytg ==8), "FG", coaches.do),
  coaches.do = ifelse((yfog > 66 & ytg ==9), "FG", coaches.do),
  coaches.do = ifelse((yfog > 66 & ytg ==10), "FG", coaches.do)
  )

df.goforit <- df.goforit %>%
  mutate(
   coaches.should = ifelse((ytg == 1), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 28 & ytg == 2), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 41 & yfog <=80 & ytg == 3), "Go for it", coaches.should), 
   coaches.should = ifelse((yfog > 80 & yfog <=96 & ytg == 3), "FG", coaches.should), 
   coaches.should = ifelse((yfog > 96 & ytg == 3), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 45 & yfog <=72 & ytg == 4), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 72 & ytg == 4), "FG", coaches.should),
   coaches.should = ifelse((yfog > 50 & yfog <=68 & ytg == 5), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 68 & ytg == 5), "FG", coaches.should),
   coaches.should = ifelse((yfog > 53 & yfog <= 67 & ytg == 6), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 67 & ytg == 6), "FG", coaches.should),
   coaches.should = ifelse((yfog > 59 & yfog <= 65 & ytg == 7), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 65 & ytg == 7), "FG", coaches.should),
   coaches.should = ifelse((yfog > 59 & yfog <=64 & ytg == 8), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 64 & ytg == 8), "FG", coaches.should),
   coaches.should = ifelse((yfog > 62 & yfog <=65 & ytg == 9), "Go for it", coaches.should),
   coaches.should = ifelse((yfog > 65 & ytg == 9), "FG", coaches.should),
   coaches.should = ifelse((yfog > 63 & ytg == 10), "FG", coaches.should)
   )

df.goforit <- df.goforit %>% 
  filter(yfog + ytg <= 100 & yfog + ytg >= 10) %>% 
  mutate(dwn = 4)

table(df.goforit$coaches.do, df.goforit$coaches.should)

```

## More filtering and cleaning

```{r}
## Data check -- do we have 8 regular season home games for each team?
NFLwin %>% 
  group_by(gid) %>% 
  slice(1) %>% 
  ungroup() %>% 
  group_by(seas, h) %>% 
  count() %>% 
  print.data.frame()

### Did the team go for it on 4th down?
NFLwin <- NFLwin %>% 
  mutate(Attempt.4 = ifelse(dwn == 4 & type == "RUSH" |  dwn == 4 & type == "PASS", 1, 0))

### Filter to 4th down plays only, re-define time as a minute
NFLwin.fourth <- NFLwin %>%
   filter(dwn==4) %>% 
   mutate(time = floor((3600 - seconds)/60))

#Weather Variables
Precipitation <- c("Chance Rain",  "Flurries", "Light Rain", 
                   "Light Showers", "Light Snow", "Rain", "Snow", "Thunderstorms")
Dome <- c("Closed Roof", "Covered Roof", "Dome")


NFLwin.fourth$cond.cat <- "DRY"
NFLwin.fourth[NFLwin.fourth$cond %in% Precipitation,]$cond.cat <- "Precipitation"
NFLwin.fourth[NFLwin.fourth$cond %in% Dome,]$cond.cat <- "Dome"

NFLwin.fourth$cond.cat <- factor(NFLwin.fourth$cond.cat, levels = c("DRY", "Dome",  "Precipitation"))

##Redefine Point Differential as a Categorical Variable and then as an ordered factor
NFLwin.fourth$pointdiff <- cut(NFLwin.fourth$ptsdiff,
                     c(-Inf, -17, -9, -4, -1, .1, 3, 8, 16, Inf),
                     labels=c("pdiff.M4", "pdiff.M3",  "pdiff.M2", "pdiff.M1", 
                              "pdiff.T", "pdiff.P1", "pdiff.P2", "pdiff.P3", "pdiff.P4"))

NFLwin.fourth$pointdiff <- factor(NFLwin.fourth$pointdiff, 
                                  levels = c("pdiff.M4", "pdiff.M3",  "pdiff.M2", "pdiff.M1", 
                                             "pdiff.T", "pdiff.P1", "pdiff.P2", "pdiff.P3", "pdiff.P4"))

## Create the final set of fourth down plays
NFL <- inner_join(NFLwin.fourth, df.goforit) %>% ungroup()
NFL <- NFL[!is.na(NFL$deltawp.off.Scrapr),] 
NFL <- NFL %>% filter(pid != 588564)  #win probability calculation is off in the Lock model for this play (no time left, 15 point game)

dim(NFL)
table(NFL$Attempt.4)
table(NFL$coaches.should)
table(NFL$Attempt.4, NFL$coaches.should)

##Check the number of plays within the go for it range within each individual game
NFL %>% 
  filter(coaches.should == "Go for it") %>% 
  group_by(gid, off) %>% 
  count() %>% 
  group_by(n) %>% 
  count()

NFL %>% 
  filter(coaches.should == "Go for it") %>% 
  group_by(gid, off) %>% 
  count() %>% 
  group_by(n) %>% 
  count() %>% 
  mutate(tot = nn*n) %>% 
  ungroup() %>% 
  summarise(AverageNumberPerGame = sum(tot)/sum(nn))


#Export the data set into a new csv file. to be used in matching
#write.csv(NFL, file = "Data/NFL_fourthdown_cleaned.csv", row.names = FALSE)
```

## Create a plot of the proportion of fourth downs attempted over the past 13 seasons.

```{r}
head(NFL)

prop.df <- NFL %>% 
  group_by(seas) %>% 
  summarise(propatt = mean(Attempt.4))

propatt.gg <- ggplot(prop.df, aes(seas, propatt)) + geom_point(size = 3) + 
  ggtitle("Proportion of fourth downs attempted within the Bot's go for it range") +
  xlab("Season") + ylab("Proportion Attempted") + 
  theme(plot.title = element_text(hjust = 0.5, size = rel(1.2)),
        axis.title.y = element_text(size = rel(1.2)),
        axis.title.x = element_text(size = rel(1.2)))

propatt.gg

#ggsave(file = "Figures/PropAttempt_per_season.png", plot = propatt.gg)

```








