---
title: "Data Wrangling"
author: "Derrick Yam & Michael Lopez"
date: "December, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Required libraries

```{r}
library(readr)
library(dplyr) 
library(splines)
library(ggplot2)
library(broom)
library(xtable)
```


##Matched Data 
```{r, include = FALSE}
games <- read_csv("Data/GAME.csv")
matched.subset.atc <- read_csv("Data/matched.subset.atc.csv")
## outcome variable 
matched.subset.atc <- matched.subset.atc %>% 
  mutate(deltawp.off.ave = (deltawp.off.Lock + deltawp.off.Scrapr)/2)
matched.subset.atc$pair.id <- rep(1:(nrow(matched.subset.atc)/2), 2)
matched.subset.atc %>% filter(pid == 596302)
```


## Overall average changes in WP

```{r}
sum.matched <- matched.subset.atc %>% 
  group_by(Attempt.4) %>% 
  summarise(mean.delta = mean(deltawp.off.ave), med.delta = median(deltawp.off.ave), n.plays = n())
sum.matched
```

## Wilcoxon Rank Sum Test ATC

```{r}
wilcox.test(deltawp.off.ave ~ Attempt.4, matched.subset.atc)
```

## Density curve -- change in win probability between each of the two groups

```{r}
#
pd <- ggplot(matched.subset.atc, aes(x = deltawp.off.ave)) + 
  geom_density(position = "identity", alpha = 0.5, aes(fill = as.factor(Attempt.4))) + 
  scale_fill_brewer(type = "qual", palette = "Dark2") + 
  ggtitle("Change in win probability after fourth down play") + 
  xlab("Change in win probability") + ylab("Density") +
  theme(legend.position = "none") + 
  xlim(-.25, .25) + 
  annotate("text", -.13, 4, label = "Went for it", colour = "#d95f02", alpha = 1) + 
  annotate("text", .07, 15, label = "Did not go for it", colour = "#1b9e77", alpha = 1) 
  
pd

#ggsave(file = "Figures/Results_Density_Curve_ave_wp.png", plot = pd)
```

## Bootstrap for wins added per team 

Done for each of the Lock win probability model and the nflscrapR win probability model

```{r}
set.seed(1)

#Lock WP Results
teams.b <- sort(unique(matched.subset.atc$off))
df.out.lock <- data.frame(nrow = 32, ncol = 5, NA)

for (j in 1:length(teams.b)) {

df0.b <- matched.subset.atc %>% 
  filter(off == teams.b[j], Attempt.4 == 0) %>% 
  select(off, wp.Lock, predict.fitms2, deltawp.off.Lock, pair.id)
#The pair is the original row number from the data set we matched from
df1.b <- matched.subset.atc %>% 
  filter(Attempt.4 == 1, pair.id %in% df0.b$pair.id) %>% 
  select(off, wp.Lock, predict.fitms2, deltawp.off.Lock)

paired.diff.b <- df1.b$deltawp.off.Lock - df0.b$deltawp.off.Lock
         
 
B <- 10^4
boot.diffwp <- numeric(B)
n.team.b <- length(paired.diff.b)
diffwp.sum.b <- NULL
for (i in 1:B)
  {
  x.matches.b<-sample(paired.diff.b, n.team.b, replace=T)
  diffwp.sum.b[i] <- sum(x.matches.b)
 }

ci.b <- quantile(diffwp.sum.b, c(0.025, 0.5, 0.975), na.rm = TRUE)

print(paste(teams.b[j], ci.b[1], ci.b[3]), quote = FALSE)
lower.bound.b <- ci.b[1]
upper.bound.b <- ci.b[3]
med.b <- ci.b[2]


df.out.lock[j,1] <- teams.b[j]
df.out.lock[j,2] <- lower.bound.b
df.out.lock[j,3] <- med.b
df.out.lock[j,4] <- upper.bound.b
df.out.lock[j,5] <- n.team.b
}

colnames(df.out.lock) <- c("team", "lower.bound", "med", "upper.bound", "n.team")

df.out.lock %>% 
  summarise(mean = mean(med), max = max(med), min = min(med))


#Scrapr WP Results
teams.b <- sort(unique(matched.subset.atc$off))
df.out.scrapr <- data.frame(nrow = 32, ncol = 5, NA)

for (j in 1:length(teams.b)) {

df0.b <- matched.subset.atc %>% 
  filter(off == teams.b[j], Attempt.4 == 0) %>% 
  select(off, wp.Scrapr, predict.fitms2, deltawp.off.Scrapr, pair.id)
#The pair is the original row number from the data set we matched from
df1.b <- matched.subset.atc %>% 
  filter(Attempt.4 == 1, pair.id %in% df0.b$pair.id) %>% 
  select(off, wp.Scrapr, predict.fitms2, deltawp.off.Scrapr)

paired.diff.b <- df1.b$deltawp.off.Scrapr - df0.b$deltawp.off.Scrapr
         
 
B <- 10^4
boot.diffwp <- numeric(B)
n.team.b <- length(paired.diff.b)
diffwp.sum.b <- NULL
for (i in 1:B)
  {
  x.matches.b<-sample(paired.diff.b, n.team.b, replace=T)
  diffwp.sum.b[i] <- sum(x.matches.b)
 }

ci.b <- quantile(diffwp.sum.b, c(0.025, 0.5, 0.975), na.rm = TRUE)

print(paste(teams.b[j], ci.b[1], ci.b[3]), quote = FALSE)
lower.bound.b <- ci.b[1]
upper.bound.b <- ci.b[3]
med.b <- ci.b[2]


df.out.scrapr[j,1] <- teams.b[j]
df.out.scrapr[j,2] <- lower.bound.b
df.out.scrapr[j,3] <- med.b
df.out.scrapr[j,4] <- upper.bound.b
df.out.scrapr[j,5] <- n.team.b
}

colnames(df.out.scrapr) <- c("team", "lower.bound", "med", "upper.bound", "n.team")

df.out.scrapr %>% 
  summarise(mean = mean(med), max = max(med), min = min(med))



df.out.lock$type <- "Lock"
df.out.scrapr$type <- "scrapR"
df.all <- bind_rows(df.out.lock, df.out.scrapr)
df.all %>% head()

df.all.ave <- df.all %>% group_by(team) %>% summarise(med.both = mean(med))
df.all <- df.all %>% left_join(df.all.ave) %>% arrange(med.both)
df.all$team.fact <- rep(1:32, each = 2) + rep(c(.1, -.1), 32)

df.all.ave <- df.all %>% group_by(team) %>% summarise(med.both = mean(med))
df.all.ave$type <- "Mean"
df.all.ave <- df.all.ave %>% arrange(med.both)
df.all.ave$team.fact <- rep(1:32)

LvSplot <- ggplot(df.all, aes(team.fact, med, colour = type)) + 
  geom_point() + 
  geom_hline(yintercept = 0, lty = 2, col = "red") +  
  geom_errorbar(aes(ymin = (lower.bound), 
                    ymax = (upper.bound))) + 
  geom_point(data = df.all.ave, aes(x= team.fact, y = med.both, colour = "Average")) +
  scale_x_continuous(name = "Franchise",
                   breaks = 1:32, 
                   labels = unique(df.all$team)) + 
  scale_y_continuous(name = "Wins Added") + 
  coord_flip()

wins.added <- LvSplot + 
  theme_bw(10) + 
  scale_colour_manual(name = "Win Probability",
                      values = c("red", "black", "grey66"), breaks = c("Lock", "Average", "scrapR")) + 
  ggtitle("Estimated Number of Wins Added from 2004 to 2016") +
  theme(legend.position = c(.85, .15))
wins.added

ggsave(wins.added, file = "Figures/Results_Scaled.png", dpi = 600, width = 5.5, height = 7, units = "in")

df.all.ave %>% 
  summarise(mean = mean(med.both), max = max(med.both), min = min(med.both))
```

## Average Delta WP Results

```{r}
set.seed(1)

teams.b <- sort(unique(matched.subset.atc$off))
df.out.ave <- data.frame(nrow = 32, ncol = 5, NA)

for (j in 1:length(teams.b)) {

df0.b <- matched.subset.atc %>% 
  filter(off == teams.b[j], Attempt.4 == 0) %>% 
  select(off, wp.ave, predict.fitms2, deltawp.off.ave, pair.id)
#The pair is the original row number from the data set we matched from
df1.b <- matched.subset.atc %>% 
  filter(Attempt.4 == 1, pair.id %in% df0.b$pair.id) %>% 
  select(off, wp.ave, predict.fitms2, deltawp.off.ave)

paired.diff.b <- df1.b$deltawp.off.ave - df0.b$deltawp.off.ave
         
 
B <- 10^4
boot.diffwp <- numeric(B)
n.team.b <- length(paired.diff.b)
diffwp.sum.b <- NULL
for (i in 1:B)
  {
  x.matches.b<-sample(paired.diff.b, n.team.b, replace=T)
  diffwp.sum.b[i] <- sum(x.matches.b)
 }

ci.b <- quantile(diffwp.sum.b, c(0.025, 0.5, 0.975), na.rm = TRUE)

print(paste(teams.b[j], ci.b[1],  ci.b[2], ci.b[3]), quote = FALSE)
lower.bound.b <- ci.b[1]
upper.bound.b <- ci.b[3]
med.b <- ci.b[2]


df.out.ave[j,1] <- teams.b[j]
df.out.ave[j,2] <- lower.bound.b
df.out.ave[j,3] <- med.b
df.out.ave[j,4] <- upper.bound.b
df.out.ave[j,5] <- n.team.b
}

colnames(df.out.ave) <- c("team", "lower.bound", "med", "upper.bound", "n.team")

df.out.ave %>% 
  summarise(mean = mean(med), max = max(med), min = min(med))

```

## Bootstrap for layered density curves plot 

```{r}
set.seed(0)
teams <- sort(unique(matched.subset.atc$off))
df.out <- NULL

for (j in 1:length(teams.b)) {

df0.b <- matched.subset.atc %>% 
  filter(off == teams.b[j], Attempt.4 == 0) %>% 
  select(off, wp.ave, predict.fitms2, deltawp.off.ave, pair.id)
#The pair is the original row number from the data set we matched from
df1.b <- matched.subset.atc %>% 
  filter(Attempt.4 == 1, pair.id %in% df0.b$pair.id) %>% 
  select(off, wp.ave, predict.fitms2, deltawp.off.ave)

paired.diff <- df1.b$deltawp.off.ave - df0.b$deltawp.off.ave
  
#Individidual bootstraps per team  
  B <- 10^4
  boot.diffwp <- numeric(B)
  n.team <- length(paired.diff)
  diffwp.sum <- NULL
  for (i in 1:B)
  {
    x.matches<-sample(paired.diff, n.team, replace=T)
    diffwp.sum[i] <- sum(x.matches)
  }
  
  df.out.temp <- data.frame(diffwp.sum, team = teams[j])
  df.out <- bind_rows(df.out, df.out.temp)
}
head(df.out)

df <- df.out %>% 
  mutate(GroupNum = rev(as.numeric(as.factor(team)))) %>% #rev() means the ordering will be from top to bottom
  group_by(team, GroupNum) %>% 
  do(tidy(density(.$diffwp.sum, bw = diff(range(.$diffwp.sum))/20))) %>% #The original has quite a large bandwidth
  group_by() %>% 
  filter(x > -4, x < 12) %>% #DRY: Range for estimated wins added?
  mutate(ymin = GroupNum * (max(y) / 1.5), #This constant controls how much overlap between groups there is
         ymax = y + ymin,
         ylabel = ymin + min(ymin)/2,
         xlabel = min(x) - mean(range(x))/5) #This constant controls how far to the left the labels are

#df.out$team<-factor(df.out$team,levels = df.out$team[order(df.out$median)])  

#Get quartiles
labels <- df.out %>% 
  mutate(GroupNum = rev(as.numeric(as.factor(team)))) %>%  #GroupNum in order to look through the teams #I think here when we make a groupnumber we should make a group number based off the median rank to put the joy division plot interms of most wins added
  group_by(team, GroupNum) %>% 
  mutate(q1 = quantile(diffwp.sum)[2],
         median = quantile(diffwp.sum)[3],
         q3 = quantile(diffwp.sum)[4]) %>%
  filter(row_number() == 1) %>% 
  dplyr::select(-diffwp.sum) %>% 
  left_join(df) %>% 
  mutate(xmed = x[which.min(abs(x - median))],
         yminmed = ymin[which.min(abs(x - median))],
         ymaxmed = ymax[which.min(abs(x - median))]) %>% 
  filter(row_number() == 1)


#df.out$team<-factor(df.out$team,levels = df.out$team[order(df.out$median)])  


p <- ggplot(df, aes(x, ymin = ymin, ymax = ymax)) + 
  geom_text(data = labels, aes(xlabel, ylabel, label = team)) +
  geom_vline(xintercept = 0, size = 1.5, alpha = 0.5, colour = "#626262") + 
  geom_vline(xintercept = c(-2, -1, 0, 1, 2, 3, 4, 5, 6, 7, 8), 
             size = 0.75, alpha = 0.25, colour = "#626262") + 
  theme(panel.grid = element_blank(),
        panel.background = element_rect(fill = "#F0F0F0"),
        axis.text.y = element_blank(),
        axis.text.x =element_text(size=15),
        axis.ticks = element_blank(),
        axis.title = element_blank()) 



#Loop to create each density curve
for (i in unique(df$GroupNum)) {
  p <- p + geom_ribbon(data = df[df$GroupNum == i,],
                       aes(group = GroupNum), #group = median rank
                       colour = "#F0F0F0", fill = "black") +
    geom_segment(data = labels[labels$GroupNum == i,],
                 aes(x = xmed, xend = xmed, y = yminmed, yend = ymaxmed),
                 colour = "#F0F0F0", linetype = "dashed") +
    geom_segment(data = labels[labels$GroupNum == i,],
                 x = min(df$x), xend = max(df$x),
                 aes(y = ymin, yend = ymin), size = 1, lineend = "round") 
}


#Adding labels to the density curves.
p <- p + geom_text(data = labels[labels$team == "ARI",], aes(xmed - xlabel/50, ylabel), 
                   label = "Median", colour = "#F0F0F0", hjust = 0, fontface = "italic", size = 3.5)  
joy.plot <- p + ggtitle("Estimated Number of Wins Added from 2004 to 2016") + xlab("Number of Wins")
joy.plot

ggsave(joy.plot, file = "Figures/Joy_Division_ave_wp.png", height = 8.35, width = 6.8)

```

## Extra plots looking at team level traits

```{r}
## Initial fourth down data (how often did teams go for it when they were supposed to?)
NFL.gfi <- read_csv("Data/NFL.gfi.csv")

df.team <- NFL.gfi %>% 
  group_by(off) %>% 
  summarise(gfi.rate = mean(Attempt.4), gfi.n = n()) %>% 
  left_join(df.all.ave, by = c("off" = "team")) %>% 
  dplyr::select(-type, -team.fact) %>% 
  rename(Team = off, wins.added = med.both)

head(df.team)

# explort to .tex
print(xtable(df.team), include.rownames=FALSE) #prints the table for latex in the console

summary(lm(wins.added ~ gfi.rate, data = df.team))
summary(lm(wins.added ~ gfi.n, data = df.team))
cor(df.team[,2:4])

### wins added versus win percentage
wins.road <- games %>% 
  filter(seas > 2003) %>%
  filter(wk < 18) %>%
  group_by(v) %>%
  summarise(win.road = sum(ptsv > ptsh))

wins.home <- games %>%
  filter(seas > 2003) %>%
  filter(wk < 18) %>%
  group_by(h) %>%
  summarise(win.home = sum(ptsv < ptsh))

df.wins <- data.frame(wins.road$v, wins.road$win.road, wins.home$win.home)

df.wins <- df.wins %>%
  rename(team = wins.road.v) %>% 
  mutate(wins = wins.road.win.road + wins.home.win.home) %>% 
  dplyr::select(team, wins)


wvwp.df2 <- left_join(df.all.ave, df.wins)
head(wvwp.df2)
cor(wvwp.df2[,c(2, 5)])
summary(lm(wins ~ med.both, data = wvwp.df2))
## note: each team plays 16 games a season, and so correlation with wins is identical to win percentage
```






